Autopolyploid establishment through polygenic adaptation

### 2025-04-07 doc/paper/response.md

I have expanded the discussion slightly to address some of the remarks of
Reviewer 2 concerning the issues of mutation and the scaling of genetic variance. 
I have added a concluding paragraph following the suggestion of Reviewer 1 to
highlight that other eco-evolutionary questions about mixed-ploidy populations
might be fruitfully addresed using the infinitesimal framework. I took the
opportunity to allude to the work of @oswald2011 in that section, following up
on a suggestion of Reviewer 2.

Overall, I think these revisions have contributed considerably to a
better-motivated article that engages more with relevant empirical work. I
hence thank the reviewers and the editors for their suggestions. Below I
include a detailed reply to both reviewers.

## Error in the first submission

In the simulations included in the first submission of my article, I did not
sample the trait values of the next generation correctly when there was nonzero
selection $\gamma > 0$ (as in figs. 2-5).
As outlined in my paper, I sample the number of offspring of a particular
combination of gametes from a particular parental pair after selection in
proportion with the expected fitness using eq. 11 (equation number in new
version).
In the previous version, I then sampled, incorrectly, offspring trait values
from the Gaussian distribution of offspring implied by the mixed-ploidy
infinitesimal model (eq. 7). This however fails to account for selection on the
trait value within each parental pair.

The density of trait values *after* selection is however also Gaussian and
found easily.
Writing $f(z)$ for the Gaussian density of trait values with mean $\bar{z}$ and
variance $V$, and $w(z)$ for the fitness function, the density after selection
$f^\ast(z)$ is
\begin{align*}
f^\ast(z) &= \frac{w(z)f(z)}{\int w(z)f(z)dz} \\
&\propto \exp\left(\gamma(z - \theta) - \frac{(z - \bar{z})^2}{2V}\right) \\
&\propto \exp\left(-\frac{1}{2V}(z^2 - 2z\bar{z} - 2V\gamma z)\right) \\
&= \exp\left(-\frac{1}{2V}\left[z^2 - 2z(\bar{z} + \gamma V) + (\bar{z} + \gamma V)^2 - (\bar{z} + \gamma V)^2\right]\right) \\
&\propto \exp\left(-\frac{1}{2V}\left[z - (\bar{z} + \gamma V)\right]^2\right)
\end{align*}
i.e. $f^\ast(z)$ is a Gaussian density with mean $\bar{z} + \gamma V$ and
variance $V$. 
The correct trait value distribution for the offspring in parental pair $(i,j)$
with gametes of ploidy $k$ and $l$ *after selection* is hence a Gaussian with
mean $\overline{z_{ij}^{kl}} + \gamma V_{ij}^{kl}$ and variance $V_{ij}^{kl}$.

The incorrect results from the previous submission can be seen as a kind of
'zeroth order' approximation to the correct results: i.e. they implement
selection on the mean trait value across families, but not selection on the
variance. The main consequence of this error is that adaptation is more
efficient than in my previous results, and hence establishment is easier. This
is however the case for both diploids and tetraploids. 
The effect on establishment is however nonlinear, so although the main qualitative
results of the paper remain the same, the new results do not simply amount to a
rescaling of the old ones.

@@

>Results are already well discussed when presented. I would either merge the
results and discussion sections (as in brief communications for example),
because it reads well as presented, or remove any explanations/discussion from
the results to put them in the discussion section. The first option sounds
better to me.

I appreciate the comment and have tried to rework the text with this in mind.
However, following the suggestion of the associate editor, I went with the
second option. I refer to my answer to the associated editor above.

>A small concluding section where the authors mentioned other questions to
study with their model will be useful in my opinion. The author proposed to
complexify the current model of adaptation with dominance or to relax some
hypotheses, but I think other biological scenarios could be studied too.

Whereas I don't think listing potential biological scenario's that can be
investigated using my modeling approach is very useful, I think the reviewer is
right in that it may be useful to highlight that the modeling approach can be
employed to address other questions related to the ecology and evolution of
mixed-ploidy populations. I have now added a small paragraph at the end that
does so (lines 549-561).

>Defining establishment when $N=100$: Is it a classic way to define
establishment in such models?  Or does it come from your experience with the
model and you see that Under $N=100$ extinction remains likely? More details
would be needed here.

@@

Of course, when considering the mixed-ploidy infinitesimal model independently
of the establishment question, mutation does become a relevant topic. I mention
this in the revised discussion but I think it is out of scope to discuss this
in any detail in the paper. I think that my approach for scaling across ploidy
levels that is explicitly based on underlying allelic effects in a finite-locus
model extends readily to mutation. The main difference between ploidy levels is
of course that the mutational target size is proportional to the ploidy level,
so that, unless the mutational variance is exactly halved in tetraploids, the
contribution of mutation to genetic variance ($\mu V_m$) in tetraploids is
larger than in diploids.

I have added a note in the discussion regarding mutation and why I have ignored it.

>In the intro providing empirical examples of polyploid establishment in
peripheral habitats would broaden readership and connections to the
literature. Also, reviewing empirical work on the genetics of adaptation to
peripheral environments in polyploids would broaden connections to the
literature.

I agree with the relevance of empirical work on polyploid establishment and
adaptation to marginal habitats for the introduction of the paper. However, I
think that to present an actual review of empirical studies here is a bit out
of scope. Griswold presents a brief review in his 2021 paper, and I prefer to
refer to his paper instead of paraphrasing his paragraph. Specifically, I now
write (lines 63-65):
"Many empirical studies of mixed-ploidy populations find that polyploids
established in peripheral habitats at the edge of a species’ range (reviewed in
Griswold (2021)), and this is in accord with large scale biogeographical
patterns (Rice et al., 2019)".

While I think it is out of scope to review in any depth the empirical
literature on adaptation and polyploidy, I agree that the paper lacked a clear
connection to such work, and in particular I noticed that it lacked a clear
motivation for the interest in *polygenic* adaptation.
I have now tried to better motivate the likely polygenic nature of local
adaptation and have added a couple of references to recent work on the genetics
of adaptation in autopolyploids (lines 75-82): 
"More often than not, local adaptation is polygenic in nature (Pritchard and Di
Rienzo, 2010; Barghi et al., 2020; Bomblies and Peichel, 2022), involving many
weakly selected variants across the genome, and adaptation during polyploid
establishment is unlikely to be an exception. Recent studies on local
adaptation in autopolyploids indeed tend to find a polygenic basis of
adaptation (Bohutı́nská et al., 2021; Konečná et al., 2021, 2022), however it is
not clear how observed adaptive differentiation in established tetraploid
populations relates to adaptation that occurred during initial establishment".

@@

>The discussion could relate findings to other theoretical works more. For
example, it does not go back and place the results in the context of Oswald and
Nuismer (2011), which assumed a two-locus model and varied dosage effects, as
well as competition between diploids and tetraploids, including no competition.
In addition it allowed for assortative mating.

I agree that @oswald2011 could be a relevant paper to relate my work to,
however it should be noted that this paper focuses on establishment in
sympatry, and does not deal with migration, nor establishment in a new
environment (they do study establishment after environmental change, but as far
as I can tell they assumed a constant population size, so this is quite
different from establishment in a new habitat).
While my study is clearly related to theirs in that it is concerned with
establishment of autotetraploids in mixed-ploidy populations, it is not
straightforward to explicitly relate the results beyond noting that (1) they
investigated more processes (assortative mating by trait value, competition,
stabilizing selection and inbreeding depression) and (2) also found that
assortative mating and selfing may promote establishment.  The latter was of
course already found by many authors before @oswald2011...

However, I have now added a more explicit reference to @oswald2011 in the concluding
note (lines 549-561, see also the associated remark of Reviewer 1), to
highlight that it would be possible to do a study that is comparable to theirs
but assuming infinitesimal quantitative genetics instead of a model with a
small number of loci.

@@

>It would be informative to expand on the the consequences of different
scalings of genetic effects. This paper made the sensible choice to scale such
that segregation variance is equal between ploidies as a baseline. Griswold
(2021) scaled genetic effects such that autotetraploid and diploid individuals
have equal fitness as a function of proportional allele count in an individual.
A consequence of Griswold’s scaling is autotetraploids have lower additive
genetic variance at HWE (Gallais 2003, p. 185), so are expected to respond more
slowly in the absence of another process. Gallais (2003, p. 186) mentions some
studies that compared additive genetic variance between diploids and
autotetraploids.

The consequences of different scalings is taken up in the results and
discussion throughout the paper where I believe this is relevant.
In the methods section the relationship between allelic effects and additive
genetic variance is stated explicitly.
In the results shown in Fig. 2, the effect of scaling assumptions on
establishment probability is explicitly shown and described.
In the section on establishment with recurrent migration, the phenomenon that
polyploid offspring is more extreme on average when allelic effects are not
exactly scaled by one half (as in Griswold) is taken up. 
I now added some further discussion, referring explicitly to Griswold's scaling
in the Discussion section (lines 462-475). I now also refer to @porturas2019
and @gallais2003.

@@

>p.7: Suggest adding “, across gene copy arrangements and ploidies” at the end
of the last sentence before the “Establishment” section.

I do not fully understand this suggestion. I personally find this makes the
sentence more confusing, as I never talk of gene copy arrangements.

>p. 7, Establishment section: The organization of paragraphs two and three in
this section is a bit confusing. N is used before it is formally defined. In my
first reading I thought there was migration and then selection, whereas there is
selection and then migration. It would also help support a self-contained paper
to give the expression for $E\left[e^{\gamma(\dots)}\right]$.

As far as I can tell, $N$ is defined when first mentioned (line 216). I
slightly altered the sentence to make this more clear.

@@

The life cycle is indeed as follows: migrants arrive, they mate randomly and
offspring survives with a probability that is determined by its fitness. This
is indeed 'selection before migration'. I tried to make this more clear (lines
218-219).

>p. 11, figure 3 (and later figure 5): The smallest value of m is 0.1 in these
figures, but based on figure 4, it looks like m = 0.01 is a transition point from
processes shaped by the accumulation of inbreeding versus migration becoming
more of a factor. Can m be extended to 0.01?

I extended the relevant figures as suggested, as $m=0.01$ is perhaps more
representative of weak migration than $m=0.1$
Weak migration means essentially that $m^2$ is negligible, so migrants and
early generation descendants thereof do not mate with migrants and the whole
process amounts more or less to independent establishment trials (there is no
swamping).

Note that I now also included lines showing the results for the limiting
establishment probabilities as $m \rightarrow 0$, i.e.
  $$\lim_{m \rightarrow 0}P_{4} = \frac{\sum_k P_{k,4}\pi_k}{\sum_{k}(P_{k,2} + P_{k,4})\pi_k}$$
where $\pi_k$ is the equilibrium frequency of ploidy level $k$ on the mainland,
and $P_{k,j}$ is the probability of $j$-ploid establishment when the initial
migrant is of ploidy level $k$.
These are more informative than the horizontal lines in the previous version,
which were a somewhat *ad hoc* extrapolation of the results from figure 2 but
not really the proper limiting probabilities.

I do not think, however, that the previous version of fig 4 suggests such a
transition point though. As noted in the main text (previous version) the
difference between the $m=0$ and $m=0.01$ results in the previous fig. 4 are
mostly due to the fact that the latter assumes a Gaussian distribution of
migrant trait values, whereas the former assumes a fixed initial migrant trait
value.
As I no longer consider this case very relevant, I replaced the $m=0$ results
in fig 4. with results from simulations where the initial migrant has a trait
value drawn from a Gaussian distribution. This makes the results more
straightforward to interpret.

### 2025-04-07 doc/paper/paper.tex
update

For a fixed degree of maladaptation $\theta$, the probability of establishment
depends in a complicated way on the strength of selection.
To see this, note that the expected number of offspring of an initial migrant
of ploidy level $k$ is $e^{-\gamma \theta + \gamma^2 k\beta_k^2 V/4}$, and the
expected trait value among its offspring will be $\gamma k \beta_k^2 V/2$.
A higher intensity of selection ($\gamma$) therefore yields a stronger effect
of initial maladaptation, but also causes a stronger response in the mean trait
value.
If the genetic variance is not constant across cytotypes (i.e. $\beta_4^2 \ne
1/2$), this response will differ for different ploidy levels.
Differences across cytotypes in how the genetic variance evolves over time as a
consequence of inbreeding will then further cause rates of adaptation to
differ, leading to different establishment probabilities even when
$\beta_4^2=1/2$.

Indeed, we find that reduced inbreeding in tetraploids substantially increases the
likelihood of tetraploid establishment relative to diploids across a large
part of the parameter space (\cref{fig:est1}A).
For the $\beta_4^2 = 1/2$ case, the establishment probability for
tetraploids can be more than ten times as high as for diploids depending on the
selection gradient ($\gamma$) and the degree of maladaptation ($\theta$)
(\cref{fig:est1}A).
As the segregation variance and initial trait value are the same across these
simulations, this is a consequence only of the reduced rate of inbreeding,
which slows down the exhaustion of the genetic variance carried by the initial
migrant individual.
Evidently, the scaling of the genetic variance across ploidy levels has a
profound effect on the establishment probability, but only when $\beta_4$ is
close to 0.5 (i.e. individual alleles have almost half the effect size in
tetraploids compared to diploids) is the benefit of the slower rate of
inbreeding in tetraploids canceled by the reduction in genetic variance 
(\cref{fig:est1}B,C).

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/fig3.pdf}
\caption{
    Establishment with recurrent migration.
    (A) Expected time until a population is established in the marginal habitat
    for increasing rates of migration and different degrees of maladaptation
    ($\theta$). Results are shown for the case with $\alpha_k = 0$ for
    $k=2,3,4$. 
    (B) Proportion of simulation replicates in which tetraploids established.
    The dots connected by solid lines show simulation results with
    $\alpha_k=0$, whereas the dots connected by dashed lines
    show simulation results with $\alpha_2=1/2, \alpha_3=1/4$ and
    $\alpha_4=1/6$ (i.e. maximum $\alpha$).
    The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$,
    where $p_2$ and $p_4$ are the establishment probabilities for a single
    diploid and a single tetraploid individual in the absence of migration (as
    in \cref{fig:est1}, but with unreduced gamete formation included). 
    The line for the $\theta=3$ case lies around 0.25 and is not shown.
    The dashed horizontal line marks the proportion of tetraploid migrants
    (i.e. the proportion of tetraploids at equilibrium in the source
    population, $\approx 0.3$\%).
    The baseline predictions (horizontal lines) are based on $5\times 10^6$
    simulation replicates.
    All other results are based on 50.000 replicate simulations.
    We assume $\gamma=0.25$ and $u=v=0.05$.
    \label{fig:estwmig}}
\end{figure}

@@

As expected, we find that the time to establishment (of a population of either
ploidy level) first decreases with increasing migration as a result of a larger
influx of potentially succesful migrants, but later increases with increasing
migration due to swamping by gene flow (\cref{fig:estwmig}A).
Importantly, the tetraploid establishment probability is considerably larger
than the expected proportion of tetraploid migrants over a large part of the
parameter range (\cref{fig:estwmig}B, black dashed line).
Furthermore, we find that for small migration rates, the probability that
tetraploids establish rather than diploids can be larger than expected based on
the establishment probabilities in the absence of migration (colored lines in
\cref{fig:estwmig}), depending on how maladapted migrant individuals are.
This suggests that maladaptive gene flow in diploids can increase the
probability of tetraploid establishment in some cases.
However, at the same time we find that the tetraploid establishment probability
declines monotonically with $m$, which highlights the negative effects of MCE.

@@

Established diploid populations are more inbred on average than established
tetraploids when migration is weak, but the difference is slight except when
there is no migration at all (\cref{fig:estwmig2}, top row).
For stronger migration ($m > 0.1$), the opposite holds.
This is a result of two interacting processes.
On the one hand, inbreeding is slower in tetraploids, so that during adaptation
and establishment from a single or limited number of outbred individuals, 
the inbreeding coefficient is expected to increase less rapidly conditional on
the event of tetraploid establishment.
On the other hand, migration mostly introduces unrelated diploids, which cross
more readily with diploids than tetraploids, reducing the average relatedness
more strongly in established diploid than in tetraploid populations.

Conditional on establishment, tetraploids have a higher trait mean than
diploids (\cref{fig:estwmig2}, bottom row).
In the absence of migration, this is a consequence of the reduced rate of
inbreeding and the resulting increased adaptive potential of tetraploids.
For weak migration, the difference in trait values between diploids and
tetraploids, conditional on eventual establishment, is limited.
This indicates the beneficial effects of migration on establishment in
diploids: migration introduces new variation on which selection can act,
counteracting the loss of genetic variance due to inbreeding.
The genetic variance contributed by migration is however negligible in
tetraploids.
When migration is strong, tetraploids have markedly larger trait values
than diploids ($m=1$ in \cref{fig:estwmig2}), showing that diploids suffer
strongly from maladaptive gene flow when the population size is low, while
tetraploids enjoy some reproductive isolation.
Furthermore, in these replicates, tetraploids tend to emerge and rise in
frequency at larger population sizes on the island, and hence tend to derive
from diploids that already experienced several generations of selection.
These neotetraploids, deriving from diploid parents with $z>0$, will have more
extreme phenotypes on average (see methods) and hence be better adapted.

@@

Importantly, we assumed no intrinsic advantage or disadvantage of polyploids in
the marginal habitat, i.e. the expected fitness of a migrant individual is the
same regardless of the ploidy level.
Differences in the likelihood of polyploid establishment are hence caused
solely by aspects of autopolyploid genetics and the barrier to gene flow
between subpopulations of different ploidy levels.
This is undoubtedly unrealistic, as both trait values and fitness will often
differ systematically across ploidy levels \citep{porturas2019}.
For instance, neopolyploids are likely to suffer intrinsic fertility issues due
to meiotic irregularities associated with multivalent formation
\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
issues with endosperm development \citep{bretagnolle1995}.

Similarly implausible is the assumption of a constant equilibrium genetic
variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
most of our results (but see \cref{fig:est1}).
Empirical data on how the genetic variance scales across ploidy levels is scant
and suggests that there is no general rule \citep{gallais2003,porturas2019}.
The meta-analysis performed by \cite{porturas2019} does indicate that trait
variance across ploidy levels is often fairly constant, so the assumption of
equal genetic variance is arguably a reasonable default.
It should be noted however that other authors have made different assumptions
on how allelic effects (and hence genetic variance) scale across ploidy levels
(in particular \cite{griswold2021}, who scaled allelic effects in a way that is
equivalent to $\beta_4 = 1/2$ in our model).
Such assumptions evidently impact the likelihood of polyploid establishment 
(\cref{fig:est1}).
More empirical data on quantitative traits in experimental or natural
mixed-ploidy populations is needed to assess whether the mixed-ploidy
infinitesimal model can adequately describe the genetics of quantitative traits
across cytotypes, and to suggest plausible values for the relevant parameters
($\alpha, \beta$).

@@

While in this study we focused on polyploid establishment in a peripheral
habitat and how this is affected by migration from a diploid source, the
mixed-ploidy infinitesimal framework could be used to address many other
eco-evolutionary questions that arise in the study of mixed-ploidy populations.
For instance, it could be of interest to develop a complicated individual-based
model along the lines of \cite{oswald2011} to study the effects of selfing,
assortative mating and competition on establishment and coexistence of
tetraploids within diploid populations, but where the focal trait that
determines fitness and assortative mating is not controlled by a few
large-effect loci, but many loci of small effect.
Similarly, our model could be straightforwardly extended to include population
regulation and stabilizing selection, which would allow us to study polyploid
establishment along an environmental gradient and the potential of
polyploidization to promote range expansions \citep{polechova2015}.

>>>>

For a fixed degree of maladaptation $\theta$, the probability of establishment
depends in a complicated way on the strength of selection.
To see this, note that the expected number of offspring of an initial migrant
of ploidy level $k$ is $e^{-\gamma \theta + \gamma^2 k\beta_k^2 V/4}$, and the
expected trait value among its offspring will be $\gamma k \beta_k^2 V/2$.
A higher intensity of selection ($\gamma$) therefore yields a stronger effect
of initial maladaptation, but also causes a stronger response in the mean trait
value.
If the genetic variance is not constant across cytotypes (i.e. $\beta_4^2 \ne
1/2$), this response will differ for different ploidy levels.
Different rates of inbreeding due to differences in ploidy level will then
further cause rates of adaptation to differ, leading to different establishment
probabilities, even when $\beta_4^2=1/2$.

Indeed, we find that reduced inbreeding in tetraploids substantially increases the
likelihood of tetraploid establishment relative to diploids across a large
part of the parameter space (\cref{fig:est1}A).
For the $\beta_4^2 = 1/2$ case, the establishment probability for
tetraploids can be more than ten times as high as for diploids depending on the
selection gradient ($\gamma$) and the degree of maladaptation ($\theta$)
(\cref{fig:est1}A).
As the segregation variance and initial trait value are the same across these
simulations, this is a consequence only of the reduced rate of inbreeding,
which slows down the exhaustion of the genetic variance carried by the initial
migrant individual.
Evidently, the scaling of the genetic variance across ploidy levels has a
profound effect on the establishment probability, but only when $\beta_4$ is
close to 0.5 (i.e. individual alleles have almost half the effect size in
tetraploids compared to diploids) is the benefit of the slower rate of
inbreeding in tetraploids canceled by the reduction in the equilibrium genetic
variance (\cref{fig:est1}B,C).

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/fig3.pdf}
\caption{
    Establishment with recurrent migration.
    (A) Expected time until a population is established in the marginal habitat
    for increasing rates of migration and different degrees of maladaptation
    ($\theta$). Results are shown for the case with $\alpha_k = 0$ for
    $k=2,3,4$. 
    (B) Proportion of simulation replicates in which tetraploids established.
    The dots connected by solid lines show simulation results with
    $\alpha_k=0$, whereas the dots connected by dashed lines
    show simulation results with $\alpha_2=1/2, \alpha_3=1/4$ and
    $\alpha_4=1/6$ (i.e. maximum $\alpha$).
    The horizontal lines mark the establishment probabilities in the
    limit as $m\rightarrow 0$ (solid lines: without double reduction; dashed
    lines: maximum $\alpha$).
    The black horizontal line marks the proportion of tetraploid migrants
    (i.e. the proportion of tetraploids at equilibrium in the source
    population, $\approx 0.3$\%).
    The baseline predictions (horizontal lines) are based on 500.000 simulation
    replicates.
    All other results are based on 100.000 replicate simulations.
    We assume $\gamma=0.25$ and $u=v=0.05$.
    \label{fig:estwmig}}
\end{figure}

@@

As expected, we find that the time to establishment (of a population of either
ploidy level) first decreases with increasing migration as a result of a larger
influx of potentially succesful migrants, but later increases with increasing
migration due to swamping by gene flow (\cref{fig:estwmig}A).
Importantly, the tetraploid establishment probability is considerably larger
than the expected proportion of tetraploid migrants over almost the entire
parameter range examined (\cref{fig:estwmig}B, black dashed line).
However, the probability of tetraploid establishment does decline monotonically
with the migration rate, showing that the negative effects of MCE on tetraploid
establishment outweigh the positive effects of reduced maladaptive gene flow in
the absence of prezygotic isolation.

@@

Established diploid populations are more inbred on average than established
tetraploids when migration is weak, but the difference is slight except when
there is no migration at all (\cref{fig:estwmig2}, top row).
For stronger migration ($m > 0.1$), the opposite holds.
This is a result of two interacting processes.
On the one hand, inbreeding is slower in tetraploids, so that during adaptation
and establishment from a single or limited number of outbred individuals, 
the inbreeding coefficient is expected to increase less rapidly.
On the other hand, migration mostly introduces unrelated diploids, which cross
more readily with diploids than tetraploids, reducing the average relatedness
more strongly in established diploid than in tetraploid populations.

Conditional on establishment, tetraploids have a higher trait mean than
diploids (\cref{fig:estwmig2}, bottom row).
In the absence of migration, this is a consequence of the reduced rate of
inbreeding and the resulting increased adaptive potential of tetraploids.
For weak migration, the difference in trait values between diploids and
tetraploids, conditional on eventual establishment, is limited.
This indicates the beneficial effects of migration on establishment in
diploids: migration introduces new variation on which selection can act,
counteracting the loss of genetic variance due to inbreeding.
The genetic variance contributed by migration is however negligible in
tetraploids.
When migration is strong, tetraploids have markedly larger trait values
than diploids ($m=1$ in \cref{fig:estwmig2}), showing that diploids suffer
strongly from maladaptive gene flow when the population size is low, while
tetraploids are much more reproductively isolated from migrants.
Furthermore, in these replicates, tetraploids tend to emerge and rise in
frequency at larger population sizes on the island, and hence tend to derive
from diploids that already experienced several generations of selection.
These neotetraploids, deriving from diploid parents with $z>0$, will have more
extreme phenotypes on average (see methods) and hence be better adapted.

@@

Importantly, we assumed no intrinsic advantage or disadvantage of polyploids in
the marginal habitat, i.e. the expected fitness of a migrant individual is the
same regardless of the ploidy level.
Differences in the likelihood of polyploid establishment are hence caused
solely by aspects of autopolyploid genetics and the barrier to gene flow
between subpopulations of different ploidy levels.
This is undoubtedly unrealistic, as both trait values and fitness will often
differ systematically across ploidy levels (see e.g. \cite{porturas2019}).
For instance, neopolyploids are likely to suffer intrinsic fertility issues due
to meiotic irregularities associated with multivalent formation
\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
issues with endosperm development \citep{bretagnolle1995}.

Similarly implausible is the assumption of a constant equilibrium genetic
variance across cytotypes ($\beta_4^2=1/2$ in our model), which we used in
most of our results (but see \cref{fig:est1}).
Empirical data on how the genetic variance scales across ploidy levels is scant
and suggests that there is no general rule \citep{gallais2003,porturas2019}.
The meta-analysis performed by \cite{porturas2019} does indicate that trait
variance across ploidy levels is often fairly constant, so the assumption of
equal genetic variance is arguably a reasonable default.
It should be noted however that other authors have made different assumptions
on how allelic effects (and hence genetic variance) scale across ploidy levels
(in particular \cite{griswold2021}, who scaled allelic effects in a way that is
equivalent to $\beta_4 = 1/2$ in our model).
Such assumptions evidently impact the likelihood of polyploid establishment 
(\cref{fig:est1}).
More empirical data on quantitative traits in experimental or natural
mixed-ploidy populations is needed to assess whether the mixed-ploidy
infinitesimal model can adequately describe the genetics of quantitative traits
across cytotypes, and to suggest plausible values for the relevant parameters
($\alpha, \beta$).

@@

While in this study we focused on polyploid establishment in a peripheral
habitat and how this is affected by migration from a diploid source, the
mixed-ploidy infinitesimal framework could be used to address many other
eco-evolutionary questions that arise in the study of mixed-ploidy populations.
For instance, it could be of interest to develop a complicated individual-based
model along the lines of \cite{oswald2011} to study the effects of selfing,
assortative mating and competition on establishment and coexistence of
tetraploids within diploid populations, but where the focal trait that
determines fitness and assortative mating is not controlled by a few
large-effect loci (as in \cite{oswald2011}), but many loci of small effect.
Similarly, our model could be straightforwardly extended to include population
regulation and stabilizing selection, which would allow us to study polyploid
establishment along an environmental gradient and the potential of
polyploidization to promote range expansions \citep{polechova2015}.

### 2025-04-07 doc/paper/titlepage.tex
update

I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
version.

>>>>

I thank two anonymous reviewers and the associate editor for their constructive
criticism that resulted in substantial improvements of the present article.
I thank Christelle Fraïsse for encouragement and feedback, as well as Quinten
Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
version.
